The \dealii Library, Version 9.2

### 2020-05-28 9.2/paper.tex
Work through the first half of the paper.

\dealii version 9.2.0 was released May XYZ, 2020.
This paper provides an
overview of the new features of this release and serves as a citable
reference for the \dealii software library version 9.2. \dealii is an
object-oriented finite element library used around the world in the
development of finite element solvers. It is available for free under the
GNU Lesser General Public License (LGPL). Downloads are available at
\url{https://www.dealii.org/} and \url{https://github.com/dealii/dealii}.

The major changes of this release are:
%
\begin{itemize}
\item xy \todo[inline]{Update once we have the subsections in Section 2;
  provide cross-references to each of these subsections}
%   \item Improved support for automatic differentiation (see
%     Section~\ref{subsec:ad}),
\end{itemize}
%
These major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item \dealii{} had decent support for solving complex-valued problems
  (e.g., ones in quantum mechanics -- like the example covered by the
  \texttt{step-58} tutorial program covered below -- or for
  time-harmonic problems) for a while already. However, there were two
  areas in which support was missing. First, the UMFPACK direct solver
  packaged with \dealii{} did not support solving complex-valued
  linear problems. This has now been addressed: UMFPACK actually can
  solve such systems, we just needed to write the appropriate
  interfaces. Second, the \texttt{DataOut} class that is responsible
  for converting nodal data into information that can then be written
  into files for visualization did not know how to deal with vector-
  and tensor-valued fields whose components are complex numbers. An
  example for this is to solve the time-harmonic version of the
  Maxwell equations that has the electric and magnetic fields as
  solution. This, too, has been addressed in this release.
\item The class \texttt{DiscreteTime} is introduced to provide a more
  consistent, more readable, and less error-prone approach to control
  time-stepping algorithms within time-dependent simulations.
  The mutating interface of this class is designed to be minimal
  to enforce a number of important programming invariants, reducing
  the possibility of mistakes in the user code.
  When time-incrementation is requested within the user code, the class makes
  sure that time increases by a non-zero step size and the current step
  number is incremented accordingly.
  Furthermore, \texttt{DiscreteTime} ensures that the final time step ends
  precisely on a predefined end time. Therefore, the final step size is
  automatically lengthened or shortened to accommodate this feature.
  In addition, the class provides useful access methods which return the step
  number $n$, the values of the simulation time corresponding to $t_{n-1}$,
  $t_n$, and $t_{n+1}$, and the step-size values $t_n - t_{n-1}$ and
  $t_{n+1} - t_n$.
\item z \todo[inline]{What else? Maybe mention the updated step-12?}
\end{itemize}
%
In addition to these changes, the changelog lists more than 200 other
features and bugfixes.

@@

This release of \dealii contains a number of large and significant changes
that will be discussed in this section.

It of course also contains a
vast number of smaller changes and added functionality; the details of these
can be found
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_0_1_and_9_1_0.html}{
in the file that lists all changes for this release}, see \cite{changes91}.

@@

By release 9.1, all triangulation classes of \texttt{deal.II} have in common that the coarse grid is shared by
all processes and the actual mesh used for computations is constructed by repeated
refinement. However, this has its limitations in industrial applications where, often, the mesh comes
from an external mesh generator in the form of a file that already contains millions
or tens of millions of cells. For such configurations, applications might exhaust
available memory already while reading the mesh on each MPI process.

The new class \texttt{parallel::fullydistributed::Triangulation} targets this issue
by distributing also the coarse grid. Such
a triangulation can be created by providing to each process a \texttt{Triangulation\-De\-scrip\-tion::Description} struct, containing
1) the relevant data to construct the local part of the coarse grid, 2) the
translation of the local coarse-cell IDs to globally unique IDs, 3) the hierarchy
of mesh refinements, and 4) the owner of the cells on the active mesh level as well
as on the multigrid levels. For the current release, triangulations is set up
this way cannot be adaptively refined after construction, which is planned to be
improved for the next release.

@@

Large-scale simulations with 304,128 cores have revealed bottlenecks in release
9.1 during initialization of a number of distributed data structures due to the usage of expensive collective operations
like \texttt{MPI\_Allgather()} and \texttt{MPI\_\allowbreak
  Alltoall()}. Typical examples are the
pre-computation of those indices of the vector entries (or
other linear index ranges) owned by
each process, which were previously stored in  an array on every process.
This information is needed to set up
the  \texttt{Utilities::MPI::Par\-ti\-ti\-oner} class.
In release 9.2, we have replaced these functions in favor of
consensus algorithms~\cite{hoefler2010scalable}, which can be
found in the namespace \texttt{Utilities::\allowbreak MPI::\allowbreak ConsensusAlgorithms} (short: \texttt{CA}).
Now, only the locally relevant information about the index ranges is
(re)computed when needed, which, for more than 100 MPI processes, uses
point-to-point communications and a single \texttt{MPI\_IBarrier()}.

@@

By replacing the collective communications during set up and removing the arrays
that contain information for each process (enabled by the application of consensus
algorithms and other modifications---a full list of modifications leading to this
improvement can be found online), we were able to significantly
improve the set up
time for large-scale simulations and to solve a Poisson problem with multigrid
with \num{2.1e12} unknowns.
Figure~\ref{fig:init_costs} compares the timings of simulations of various problem
sizes (including set up) on 49,152 MPI ranks using a matrix-free solver according
to~\cite{KronbichlerKormann2019,KronbichlerWall2018} discontinuous elements of
degree $5$ in a geometric multigrid (GMG) scheme with the previous release 9.1
and with the current release 9.2. While the scaling for the V-cycle had been
very good before, many initialization routines have been considerably
improved, especially the enumeration of unknowns on the multigrid levels and
the setup of the multigrid transfer.

@@

\begin{figure}
  \begin{tikzpicture}
    \begin{loglogaxis}[
      width=0.51\textwidth,
      height=0.38\textwidth,
          xlabel={\# DoFs},
          ylabel={time [s]},
          title={9.1 release},
          xtick={1.4e7,5.7e7,2.3e8,9.1e8,3.6e9,1.4e10,5.8e10,2.3e11},
          xticklabels={1.4$\times 10^7$,5.7$\times 10^7$,2.3$\times 10^8$,9.1$\times 10^8$,3.6$\times 10^9$,1.4$\times 10^{10}$,5.8$\times 10^{10}$,2.3$\times 10^{11}$},
          tick label style={font=\scriptsize},
          xticklabel style={rotate=25,anchor=north east},
          label style={font=\scriptsize},
          legend style={font=\scriptsize},
          %legend to name=legend:scalingMG,
          %legend cell align=left,
          %legend columns = 3,
          ymin=2e-3,ymax=8e1,
          grid, thick]
          \addplot table[x index = 0, y index={1}] {\tableinitold};
          \addplot[red,mark=square*] table[x index = 0, y index={2}] {\tableinitold};
          \addplot[gnuplot@green,mark=otimes] table[x index = 0, y index={3}] {\tableinitold};
          \addplot[gnuplot@lightblue,mark=triangle] table[x index = 0, y index={4}] {\tableinitold};
          \addplot[gnuplot@purple,mark=diamond] table[x index = 0, y index={5}] {\tableinitold};
          \addplot[black,mark=x] table[x index = 0, y index={6}] {\tableinitold};
    \end{loglogaxis}
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}
    \begin{loglogaxis}[
      width=0.51\textwidth,
      height=0.38\textwidth,
          xlabel={\# DoFs},
          ylabel={time [s]},
          title={9.2 release},
          xtick={1.4e7,5.7e7,2.3e8,9.1e8,3.6e9,1.4e10,5.8e10,2.3e11},
          xticklabels={1.4$\times 10^7$,5.7$\times 10^7$,2.3$\times 10^8$,9.1$\times 10^8$,3.6$\times 10^9$,1.4$\times 10^{10}$,5.8$\times 10^{10}$,2.3$\times 10^{11}$},
          tick label style={font=\scriptsize},
          xticklabel style={rotate=25,anchor=north east}, thick,
          label style={font=\scriptsize},
          legend style={font=\scriptsize},
          legend to name=legend:init,
          legend cell align=left,
          legend columns = 6,
          ymin=2e-3,ymax=8e1,
          grid]
          \addplot table[x index = 0, y index={1}] {\tableinit};
          \addlegendentry{create mesh};
          \addplot[red,mark=square*] table[x index = 0, y index={2}] {\tableinit};
          \addlegendentry{enumerate (mg) dofs};
          \addplot[gnuplot@green,mark=otimes] table[x index = 0, y index={3}] {\tableinit};
          \addlegendentry{init matrix-free};
          \addplot[gnuplot@lightblue,mark=triangle] table[x index = 0, y index={4}] {\tableinit};
          \addlegendentry{init GMG transfer};
          \addplot[gnuplot@purple,mark=diamond] table[x index = 0, y index={5}] {\tableinit};
          \addlegendentry{init smoother};
          \addplot[black,mark=x] table[x index = 0, y index={6}] {\tableinit};
          \addlegendentry{GMG V-cycle};
    \end{loglogaxis}
  \end{tikzpicture}
  \\
  \strut\hfill\pgfplotslegendfromname{legend:init}\hfill\strut
  \caption{Comparison of initialization costs of various data structures in the 9.1 release (left) and the new 9.2 release (right) when run on 49,152 MPI ranks.}
  \label{fig:init_costs}
\end{figure}

The new code has been also applied to solve problems with adaptively refined
meshes with more than \num{4e9} unknowns. {\color{red}TODO[Timo]}

\todo[inline]{We did solve adaptive problems with more than 4B unknowns before, but some were broken for a while during this release stage due to the consensus algorithms. But we did not test systems or block vectors before. But we should not make it sound too negative.}

@@

With the previous release, \dealii supports $hp$-adaptive finite element methods
on distributed memory systems \cite{dealII91}. We implemented the bare functionality
for $hp$-adaptive methods with the objective to offer the greatest flexibility in
their application. Here, reference finite elements still had to be assigned manually
to each cell, which may not lead to an optimal mesh and is tedious.

@@

The interface is now as simple to use as the one for pure adaptive mesh refinement.
Consider the following (incomplete) listing as an example: We estimate both error and
smoothness of the finite element approximation. Further, we flag certain fractions of
cells with the highest and lowest errors for refinement and coarsening, respectively
(here: 30\% / 3\%). From those cells listed for adaptation, we specify a subset of them
for $h$- and $p$-adaptation (here: 10\% / 90\%).
\begin{c++}
Vector<float> estimated_error_per_cell (n_active_cells);
KellyErrorEstimator::estimate(
  dof_handler, ..., solution, estimated_error_per_cell, ...);
GridRefinement::refine_and_coarsen_fixed_number(
  triangulation, estimated_error_per_cell, 0.3, 0.03);

@@

The former relies on knowing an estimate for the upper error bound \cite[Thm.~3.4]{BabuskaSuri1990}.
For successive refinements, we can predict how the error will change on basis of
current error estimates and adaptation flags. In the next refinement cycle, these
predicted error estimates allow us to decide whether the choice of adaptation in
the previous cycle was justified, and provide the foundation for it in the next
one \cite{MelenkWohlmuth2001}.

>>>>

\dealii version 9.2.0 was released May XYZ, 2020.
\todo[inline]{Need to update date.}
This paper provides an
overview of the new features of this release and serves as a citable
reference for the \dealii software library version 9.2. \dealii is an
object-oriented finite element library used around the world in the
development of finite element solvers. It is available for free under the
GNU Lesser General Public License (LGPL). Downloads are available at
\url{https://www.dealii.org/} and \url{https://github.com/dealii/dealii}.

The major changes of this release are:
%
\begin{itemize}
\item xy \todo[inline]{Update once we have the subsections in Section 2;
  provide cross-references to each of these subsections}
%   \item Improved support for automatic differentiation (see
%     Section~\ref{subsec:ad}),
\end{itemize}
%
These major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item \dealii{} had decent support for solving complex-valued problems
  for a while already
  (e.g., ones in quantum mechanics -- like the equation used in the
  \texttt{step-58} tutorial program covered below -- or for
  time-harmonic problems). However, there were two
  areas in which support was missing. First, the UMFPACK direct solver
  packaged with \dealii{} did not support solving complex-valued
  linear problems. This has now been addressed: UMFPACK actually can
  solve such systems, we just needed to write the appropriate
  interfaces. Second, the \texttt{DataOut} class that is responsible
  for converting nodal data into information that can then be written
  into files for visualization did not know how to deal with vector-
  and tensor-valued fields whose components are complex numbers. An
  example for this is to solve the time-harmonic version of the
  Maxwell equations that has the electric and magnetic fields as
  solution. This, too, has been addressed in this release.
\item The new \texttt{DiscreteTime} class provides a more
  consistent, more readable, and less error-prone approach to control
  time-stepping algorithms within time-dependent simulations.
  The interface of this class is designed to be minimal
  to enforce a number of important programming invariants, reducing
  the possibility of mistakes in the user code.
  An example is that \texttt{DiscreteTime} ensures that the final time step ends
  precisely on a predefined end time, automatically
  lengthening or shortening to final time step.
\item z \todo[inline]{What else? Maybe mention the updated step-12?}
\end{itemize}
%
The changelog lists more than 240 other
features and bugfixes.

@@

This release of \dealii contains a number of large and significant changes
that will be discussed in this section.
It of course also contains a
vast number of smaller changes and added functionality; the details of these
can be found
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_0_1_and_9_1_0.html}{
in the file that lists all changes for this release}, see \cite{changes91}.

@@

Previously, all triangulation classes of \texttt{deal.II} had in common that the coarse grid is replicated by
all processes in a parallel environment, and the actual mesh used for computations is constructed by repeated
refinement. However, this has its limitations in industrial
applications, where frequently the mesh comes
from an external mesh generator in the form of a file that already contains millions
or tens of millions of cells. For such configurations, applications might exhaust
available memory already while reading the mesh on each MPI process.

The new \texttt{parallel::fullydistributed::Triangulation} class targets this issue
by distributing also the coarse grid. Such
a triangulation can be created by providing to each process a \texttt{Triangulation\-De\-scrip\-tion::Description} struct, containing
(i) the relevant data to construct the local part of the coarse grid, (ii) the
translation of the local coarse-cell IDs to globally unique IDs, (iii) the hierarchy
of mesh refinement steps, and (iv) the owner of the cells on the active mesh level as well
as on the multigrid levels. For the current release, triangulations set up
this way cannot be adaptively refined after construction, though we plann to
improve this for the next release.

@@

Large-scale simulations with up to 304,128 cores have revealed bottlenecks in release
9.1 during initialization of a number of distributed data structures, due to the usage of expensive collective operations
like \texttt{MPI\_Allgather()} and \texttt{MPI\_\allowbreak
  Alltoall()}. Typical examples are the
pre-computation of the indices of those vector entries (or
other linear index ranges) owned by
each process, which were previously stored in  an array on every process.
This information is needed to set up
the  \texttt{Utilities::MPI::Par\-ti\-ti\-oner} class.
In release 9.2, we have replaced these functions in favor of
consensus algorithms~\cite{hoefler2010scalable}, which can be
found in the namespace \texttt{Utilities::\allowbreak MPI::\allowbreak ConsensusAlgorithms} (short: \texttt{CA}).
Now, only the locally relevant information about the index ranges is
(re)computed when needed, which, for more than 100 MPI processes, uses
point-to-point communications and a single \texttt{MPI\_IBarrier()}.

@@

By replacing the collective communications during set up and removing the arrays
that contain information for each process (enabled by the application of consensus
algorithms and other modifications---a full list of modifications leading to this
improvement can be found online), we were able to significantly
improve the set up
time for large-scale simulations and to solve a Poisson problem with multigrid
with \num{2.1e12} unknowns.
Figure~\ref{fig:init_costs} compares timings of simulations of various problem
sizes (including set up) on 49,152 MPI ranks using a matrix-free
solver~\cite{KronbichlerKormann2019,KronbichlerWall2018}; this solver uses discontinuous elements of
degree $5$ in a geometric multigrid (GMG) scheme. The comparison between the previous release 9.1
and the current release 9.2 shows that while the scaling for the V-cycle had been
very good before, many initialization routines have been considerably
improved, especially the enumeration of unknowns on the multigrid levels and
the setup of the multigrid transfer.

@@

\begin{figure}
  \begin{tikzpicture}
    \begin{loglogaxis}[
      width=0.51\textwidth,
      height=0.38\textwidth,
          xlabel={\# DoFs},
          ylabel={time [s]},
          title={9.1 release},
          xtick={1.4e7,5.7e7,2.3e8,9.1e8,3.6e9,1.4e10,5.8e10,2.3e11},
          xticklabels={1.4$\times 10^7$,5.7$\times 10^7$,2.3$\times 10^8$,9.1$\times 10^8$,3.6$\times 10^9$,1.4$\times 10^{10}$,5.8$\times 10^{10}$,2.3$\times 10^{11}$},
          tick label style={font=\scriptsize},
          xticklabel style={rotate=25,anchor=north east},
          label style={font=\scriptsize},
          legend style={font=\scriptsize},
          %legend to name=legend:scalingMG,
          %legend cell align=left,
          %legend columns = 3,
          ymin=2e-3,ymax=8e1,
          grid, thick]
          \addplot table[x index = 0, y index={1}] {\tableinitold};
          \addplot[red,mark=square*] table[x index = 0, y index={2}] {\tableinitold};
          \addplot[gnuplot@green,mark=otimes] table[x index = 0, y index={3}] {\tableinitold};
          \addplot[gnuplot@lightblue,mark=triangle] table[x index = 0, y index={4}] {\tableinitold};
          \addplot[gnuplot@purple,mark=diamond] table[x index = 0, y index={5}] {\tableinitold};
          \addplot[black,mark=x] table[x index = 0, y index={6}] {\tableinitold};
    \end{loglogaxis}
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}
    \begin{loglogaxis}[
      width=0.51\textwidth,
      height=0.38\textwidth,
          xlabel={\# DoFs},
          ylabel={time [s]},
          title={9.2 release},
          xtick={1.4e7,5.7e7,2.3e8,9.1e8,3.6e9,1.4e10,5.8e10,2.3e11},
          xticklabels={1.4$\times 10^7$,5.7$\times 10^7$,2.3$\times 10^8$,9.1$\times 10^8$,3.6$\times 10^9$,1.4$\times 10^{10}$,5.8$\times 10^{10}$,2.3$\times 10^{11}$},
          tick label style={font=\scriptsize},
          xticklabel style={rotate=25,anchor=north east}, thick,
          label style={font=\scriptsize},
          legend style={font=\scriptsize},
          legend to name=legend:init,
          legend cell align=left,
          legend columns = 6,
          ymin=2e-3,ymax=8e1,
          grid]
          \addplot table[x index = 0, y index={1}] {\tableinit};
          \addlegendentry{create mesh};
          \addplot[red,mark=square*] table[x index = 0, y index={2}] {\tableinit};
          \addlegendentry{enumerate (mg) dofs};
          \addplot[gnuplot@green,mark=otimes] table[x index = 0, y index={3}] {\tableinit};
          \addlegendentry{init matrix-free};
          \addplot[gnuplot@lightblue,mark=triangle] table[x index = 0, y index={4}] {\tableinit};
          \addlegendentry{init GMG transfer};
          \addplot[gnuplot@purple,mark=diamond] table[x index = 0, y index={5}] {\tableinit};
          \addlegendentry{init smoother};
          \addplot[black,mark=x] table[x index = 0, y index={6}] {\tableinit};
          \addlegendentry{GMG V-cycle};
    \end{loglogaxis}
  \end{tikzpicture}
  \\
  \strut\hfill\pgfplotslegendfromname{legend:init}\hfill\strut
  \caption{\it Comparison of initialization costs of various data structures in the 9.1 release (left) and the new 9.2 release (right) when run on 49,152 MPI ranks.}
  \label{fig:init_costs}
\end{figure}

\todo[inline]{Timo: Do you want to say something about that the new code has been also applied to solve problems with adaptively refined
meshes with more than \num{4e9} unknowns?}

\todo[inline]{Timo says: We did solve adaptive problems with more than
  4B unknowns before, but some were broken for a while during this
  release stage due to the consensus algorithms. But we did not test
  systems or block vectors before. But we should not make it sound too
  negative.}

\todo[inline]{Wolfgang says: Should we just not mention any of this at
  all? The section stands well on its own.}

@@

Since the previous release, \dealii has had support for $hp$-adaptive finite element methods
on distributed memory systems \cite{dealII91}. We implemented the bare functionality
for $hp$-adaptive methods with the objective to offer the greatest flexibility in
their application. Here, reference finite elements still had to be assigned manually
to each cell, which may not lead to an optimal mesh and is tedious.

@@

The interface is now as simple to use as the one for $h$-adaptive mesh refinement.
Consider the following (incomplete) listing as an example: We estimate both error and
smoothness of the finite element approximation. Further, we flag certain fractions of
cells with the highest and lowest errors for refinement and coarsening, respectively
(here: 30\%/3\%). From those cells listed for adaptation, we designate a subset
for $h$- and $p$-adaptation (here: 10\%/90\%).
\begin{c++}
Vector<float> estimated_error_per_cell (n_active_cells);
KellyErrorEstimator::estimate(
  dof_handler, ..., solution, estimated_error_per_cell, ...);
GridRefinement::refine_and_coarsen_fixed_number(
  triangulation, estimated_error_per_cell, 0.3, 0.03);

@@

The former relies on knowing an estimate for the upper error bound \cite[Thm.~3.4]{BabuskaSuri1990}.
For successive refinements, we can predict how the error will change
based on
current error estimates and adaptation flags. In the next refinement cycle, these
predicted error estimates allow us to decide whether the choice of adaptation in
the previous cycle was justified, and provide a criterion for the choice in the next
cycle \cite{MelenkWohlmuth2001}.
